Text embedding models yield high resolution insights into conceptual knowledge from short multiple choice quizzes

### 2023-02-01 paper/main.tex
finished (i think! 🤞) addressing PCF's comments in slack and in the latex notes.

\subsubsection*{Constructing text embeddings of multiple lectures and questions}\label{subsec:topic-modeling}
We adapted an approach we developed in prior work~\citep{HeusEtal21} to embed
each moment of the two lectures and each question in our pool in a common representational space.
Briefly, our approach uses a topic model (Latent Dirichlet Allocation;~\citealp{BleiEtal03}), trained on a set
of documents, to discover a set of $k$ ``topics'' or ``themes.'' Formally, each
topic is defined as a set of weights over each word in the model's vocabulary
(i.e., the union of all unique words, across all documents, excluding ``stop
words.''). Conceptually, each topic is intended to give larger weights to words
that are conceptually related or that tend to co-occur in the same documents.
After fitting a topic model, each document in the training set, or any
\textit{new} document that contains at least some of the words in the model's
vocabulary, may be represented as a $k$-dimensional vector describing how much
the document (most probably) reflects each topic. (Unless, otherwise noted, we
used $k = 15$ topics.)

@@

An important feature of our approach is that, given a trained text embedding
model and participants' quiz performance on each question, we can estimate
their knowledge about \textit{any} content expressible by the embedding
model---not solely the content explicitly probed by the quiz questions or even
appearing in the lectures. To visualize these estimates
(Figs.~\ref{fig:knowledge-maps},~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,~\individualKnowledgeMapsC,~\individualLearningMapsA,
and~\individualLearningMapsB), we used Uniform Manifold Approximation and
Projection (UMAP; \citealp{McInEtal18a}) to construct a 2D projection of the
text embedding space. Sampling the original 100-dimensional space at high
resolution to obtain an adequate set of topic vectors spanning the embedding
space would be computationally intractable. However, sampling a 2D grid is much
more feasible. We defined a rectangle enclosing the 2D projections of the
lectures' and quizzes' embeddings, and we sampled points from a regular $100
\times 100$ grid of coordinates that evenly tiled the enclosing rectangle. We
sought to estimate participants' knowledge (and learning, i.e., changes in
knowledge) at each of the resulting 10,000 coordinates.

>>>>

\subsubsection*{Constructing text embeddings of multiple lectures and questions}\label{subsec:topic-modeling}

We adapted an approach we developed in prior work~\citep{HeusEtal21} to embed
each moment of the two lectures and each question in our pool in a common
representational space. Briefly, our approach uses a topic model (Latent
Dirichlet Allocation;~\citealp{BleiEtal03}), trained on a set of documents, to
discover a set of $k$ ``topics'' or ``themes.'' Formally, each topic is defined
as a set of weights over each word in the model's vocabulary (i.e., the union
of all unique words, across all documents, excluding ``stop words.'').
Conceptually, each topic is intended to give larger weights to words that are
conceptually related or that tend to co-occur in the same documents. After
fitting a topic model, each document in the training set, or any \textit{new}
document that contains at least some of the words in the model's vocabulary,
may be represented as a $k$-dimensional vector describing how much the document
(most probably) reflects each topic. (Unless, otherwise noted, we used $k = 15$
topics.)

@@

An important feature of our approach is that, given a trained text embedding
model and participants' quiz performance on each question, we can estimate
their knowledge about \textit{any} content expressible by the embedding
model---not solely the content explicitly probed by the quiz questions or even
appearing in the lectures. To visualize these estimates
(Figs.~\ref{fig:knowledge-maps},~\individualKnowledgeMapsA,~\individualKnowledgeMapsB,~\individualKnowledgeMapsC,~\individualLearningMapsA,
and~\individualLearningMapsB), we used Uniform Manifold Approximation and
Projection (UMAP; \citealp{McInEtal18a}) to construct a 2D projection of the
text embedding space. Sampling the original 100-dimensional space at high
resolution to obtain an adequate set of topic vectors spanning the embedding
space would be computationally intractable. However, sampling a 2D grid is
trivial.

At a high level, the UMAP algorithm obtains low-dimensional embeddings by
minimizing the cross entropy between the pairwise (clustered) distances between
the observations in their original (e.g., 100-dimensional) space and the
pairwise (clustered) distances in the low-dimensional embedding space (in our
approach, the embedding space is 2D). In our implementation, pairwise distances
in the original high-dimensional space were defined as 1 minus the correlation
between the pair of coordinates, and pairwise distances in the low-dimensional
embedding space were defiend as the Euclidean distance between the pair of
coordinates.

Although UMAP is not fully invertible, due to its clustering step, it is
possible to approximately invert the embedded low-dimensional points. In our
application, we sought to embed topic vectors, whose elements are always
non-negative. Because the inversion step is inexact, sometimes it results in
negative-valued vectors (even if the original observations were all
non-negative), which are incompatable with the topic modeling framework. To
protect against this issue, we log-transformed the topic vectors prior to
embedding them in the 2D space. When we inverted the embedded vectors (e.g., to
estimate topic vectors or word clouds, as in Fig.~\ref{fig:knowledge-maps}C),
we passed to inverted (log-transformed) values through the exponential function
to obtain a vector of non-negative values.

After embedding both lectures' topic trajectories and the topic vectors of
every question, we defined a rectangle enclosing the 2D projections of the
lectures' and quizzes' embeddings. We then sampled points from a regular $100
\times 100$ grid of coordinates that evenly tiled this enclosing rectangle. We
sought to estimate participants' knowledge (and learning, i.e., changes in
knowledge) at each of the resulting 10K coordinates.
